Measurement Bias from Address Aliasing

### 2014-05-18 paper.tex
Improve abstract
And other small fixes...

    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchitectures. 
    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.

    We analyze a small program that was presented as a curiosity in related previous work on \emph{measurement bias};
    Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.

    Using hardware performance counters, we identify a strong correlation with resource stalls, ultimately caused by address aliasing events.
    The CPU uses an optimistic strategy when issuing operations out of order.
    Only the last 12 address bits are compared to determine if there was a conflict, sometimes resulting in performance penalties from false dependencies.

    We show how address aliasing can cause measurement bias for two particular external properties; environment variable size, and dynamically linked heap allocation libraries.
    Performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give worst case performance by default for large allocations.

    \end{abstract}
    \bigskip

@@

\section{Introduction}
% - Measurement bias in systems research
%   - What is measurement bias
%     - fundamentally anything external
%     - Room temperature
%   - Studied before
%     - External properties that affect memory layout
%       - link order (not really)
%       - env size
%   - Has been shown to be significant and commonplace
%     - But difficult to deal with
%     - Need to apply statistical methods to experiments
%       - randomization of experimental setups
% - Want to look at flipside: understand what why bias occurs
%   - Use for optimization
%   - Find interesting cases
%   - Known causes: cache alignment, ignore.
% - Analysis: 
%   - Use known example that exhibit bias
%   - Find lesser known source in address aliasing
% - Results:
%   - Applicable general knowledge about Intel architecture
%   - Useful for optimization
%   - Libraries and compilers fail to account for it in reality.

How fast does an algorithm run?
Is program A faster than program B on \emph{my} machine?
These types of questions are, perhaps surprisingly, not always trivial to answer.
The usual approach to quantify performance is to simply use some metric, such as the wall time or number of cycles executed.
However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.

@@

@@

% Detailed comparison between median and worst cases.
\begin{table}[t]
  \caption{Events with significant correlation to cycle count.}
  \label{tab:loopcorrelation}
  \centering
  \renewcommand{\tabcolsep}{1pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\scriptsize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Median, [index]3, [index]4},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & Median & Spike 1 & Spike 2 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/micro-kernel-comparison.csv}
\end{table}

>>>>

    Seemingly innocious properties of the environment, such as the contents of system environment variables, can affect the performance of computer programs.
    This effect is known as \emph{measurement bias}, and has been shown to be significant, but unpredictable and difficult to deal with.

    In this paper we identify one of the underlying mechanics resulting in measurement bias on modern Intel microarchitectures.
    Using hardware performance counters, we identify that bias is ultimately caused by \emph{address aliasing}.
    When issuing load operations, the CPU is only comparing the last 12 address bits to previous stores, sometimes resulting in performance penalties from \emph{false} dependencies.

    % In addition to environment variable size, we show how address aliasing can cause measurement bias from variations in dynamically linked heap allocation libraries.
    We find that performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give \emph{worst case} performance by default for large allocations.
    An understanding of address aliasing is necessary in order to achieve optimal performance, and we outline how application developers can utilize this knowledge.
    \end{abstract}
    \bigskip

@@

\section{Introduction}
How fast does an algorithm run?
Is program A faster than program B on \emph{my} machine?
These types of questions are, perhaps surprisingly, not always trivial to answer.
The usual approach to quantify performance is to simply use some metric, such as the wall time or number of cycles executed.
However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.

@@

% Detailed comparison between median and worst cases.
\begin{table}[t]
  \caption{Events with significant correlation to cycle count.}
  \label{tab:loopcorrelation}
  %\centering
  \renewcommand{\tabcolsep}{1pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\scriptsize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Median, [index]3, [index]4},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & Median & Spike 1 & Spike 2 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/micro-kernel-comparison.csv}
\end{table}

@@
